---
title: "Class 3 Exercise"
author: "Swati Singh"
date: "2/26/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Scenario

You work as a data scientist at a company that sells widgets. The CEO and owner is extremely engaged in looking at the most recent data on sales but is not a statistician and is prone to pay too much attention to meaningless day-to-day and month-to-month fluctuations.

January 2, 8 AM:  The CEO comes into your office and expresses worry about widget sales for the most recent month, December.  She thinks sales have tanked and wants you to look into the situation further and provide a brief report by noon. Widget demand is seasonal and the business depends on strong holiday sales. She wants a brief report on her desk by noon. 

## Topics

1. EDA workflow
2. Practice data manipulation and visualization
3. Statistical inference
4. Communication


## Process

- Please gather into groups to work on this project!
- Put group members' names up above in the yaml heading under "author" (where it currently says "Names of those in your group").
- Collaborate on one document.
- When you are done, compile to HTML (or PDF), and submit through Canvas.

## EDA workflow

1. Formulate a question:  Are last month's sales (month 12 of year 5) down?

2. Read in your data.  For this exercise we will simulate a dataset.

```{r}

widget <- expand.grid(year = 1:5, month = 1:12, day = 1:30)

# The expand.grid() function creates a dataframe with unique combinations of the values
# from each variable. Here we are setting up a data.frame for 5 years of data.

head(widget)

set.seed(1126) # Use set.seed() to ensure identical datasets

# Now, simulate sales data using the uniform distribution and the normal distribution:
# runif() and rnorm().  For simplicity we will pretend each month has exactly
# 30 days.

widget %<>%
  mutate(sales = ifelse(month < 12, 
                        runif(5 * 11 * 30, min = 800, max = 1200) +
                          rnorm(5 * 11 * 30, mean = 100, sd = 100),
                        runif(5 * 11 * 30, min = 1100, max = 1300) +
                          rnorm(5 * 11 * 30, mean = 100, sd = 100)),
         sales = ifelse(month == 12 & year == 5, 
                        sales - rnorm(30, mean = 100, sd = 30),
                        sales),
         year = factor(year),
         month = factor(month)) %>% 
  arrange(year, month, day) %>%
  mutate(instance = 1:(5*12*30)) # instance is a row counter.

head(widget)

```

3. Check the packaging: dim(), nrow(), ncol()

```{r}
dim(widget)
nrow(widget)
ncol(widget)
```

4. Inspect the dataset: str(), glimpse(), View()

```{r}
str(widget)

glimpse(widget)

view(widget)
```

5. Look at the top and the bottom of your data: head(), tail()

```{r}
head(widget)
tail(widget)
```

6. Summarize the data: summary(), table(), hist()

```{r}
summary(widget)
numericcol<-names(select_if(widget,is.numeric))
numericcol
hist(widget$sales)
```

7. Try the easy solution first 

    + Plot daily sales
    
```{r}

ggplot(widget,aes(sales))+
  geom_density()+
  labs(title="Distribution of daily sales")

ggplot(widget,aes(instance,sales))+
  geom_line()+
  labs(title="Distribution of daily sales")
```

    + Plot monthly sales
    
    ```{r}
widget %>%
      group_by(month) %>%
      summarise(sales_over_month=sum(sales)) %>%
      ggplot(aes(month,sales_over_month,group=1))+
      geom_line() +
      labs(title="Total sales per month")

    #Alternatively    
ggplot(widget,aes(month,sales,group=1))+
      stat_summary(fun.y = sum,geom="line")
    
```
Widget sales are **much** higher in December

    + Plot yearly sales
    
    ```{r}
ggplot(widget,aes(year,sales,group=1))+
      stat_summary(fun.y = sum,geom="line")
```
   The sales in last year dropped drastically as compared to previous months
   
    + Summarize total and average sales by month, and calculate confidence intervals.

```{r}
widget %>%
  group_by(month) %>%
  summarize(total_sales=sum(sales),
            avg_sales=mean(sales),
            n=n(),
            SEM=sd(sales),
            lower=avg_sales-2*SEM,
            upper=avg_sales+2*SEM)
```
    
    + Summarize total and average sales by  year, and calculate confidence intervals. For the CIs remember that SEM = s / sqrt(n) and that the interval is mean -/+ 1.96 * SEM.

```{r}
widget %>%
  group_by(year) %>%
  summarize(total_sales=sum(sales),
            avg_sales=mean(sales),
            n=n(),
            SEM=sd(sales)/sqrt(n),
            lower=avg_sales-1.96*SEM,
            upper=avg_sales+1.96*SEM)
```

8. Challenge your solution

    + Plot monthy sales facetted by year

```{r}
ggplot(widget,aes(month,sales,group=1))+
  stat_summary(fun.y=sum,geom="line")+
  facet_wrap(~year)


```

    + Plot monthly sales colored by year


```{r}
ggplot(widget,aes(month,sales,group=year,color=year))+
  stat_summary(fun.y=sum,geom="line")
  
widget %>%
  group_by(month,year) %>%
  summarize(sum_sales=sum(sales))%>%
  ggplot(aes(month,sum_sales,group=year,color=year))+
  geom_line()+
  labs(title="monthly sales by year")
```

    + Summarize total and average sales by month and year, and calculate confidence intervals.


```{r}
widget %>%
  group_by(month,year) %>%
  summarize(total_sales=sum(sales),
            avg_sales=mean(sales),
            n=n(),
            SEM=sd(sales)/sqrt(n),
            lower=avg_sales-2*SEM,
            upper=avg_sales+2*SEM)  %>%
            filter(month==12)
```

9. Follow up questions. See next section.

## Statistical inference 

Answer the CEO's question:  Is there a real drop in year 5 month 12 sales or is the difference just due to random variation? 

How would you approach this question?


```{r}
ggplot(subset(widget,month==12),aes(month,sales,color=year))+ geom_boxplot()+
  stat_summary(fun.y=mean,geom="point")
```

## Communication 

Write a paragraph summary of your descriptive and inferential findings.

On comparing the yearly sales for month of December, we can see that the sales have much dropped as cmpared to previous years. Plotting the box plot we can see the difference between the groups and can see their mean and lower and upper boundaries are not the same. hence we can reject the null hypothesis. Now we will look at the statistical importance of the month and year for month 12 and year 2015.
We see while month plays an significant impact on the sales, the year has no significant impact.

```{r}
model<-lm(sales~month*year,widget)
summary(model)
```

